Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-21 dictionary.tex
commit before pulling from sage cloud

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $\sigma_\mathcal{L}) \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $S_G(\mathcal{L}) = S_G(\mathcal{L'})$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
S_G: \QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $S_G(\qcs{L})$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.
It is easy to see that the function $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$ is a group homomorphism.\todo{I still need to write down the details, for ourselves.}

\begin{definition}\label{def:invisible}
Let $\qcs{L}$ be a quasicharacter sheaf on $G$. We say $\qcs{L}$ is \emph{visible} $g_\qcs{L} \in H^0(\Weil{},H^2(G(\bFq),\EEx))$ is trivial; otherwise, we say $\qcs{L}$ is \emph{invisible}.
\end{definition}

%If $G$ is a connected and of finite type, then $\Hh^2(G(\bFq),\EEx)$ is trivial\todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible; by contrast, if $G$ is etale then $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$ may be non-trivial so $G$ may admit invisible quasicharacters sheaves. We will return to that case in Section~\ref{sec:etale}.

>>>>

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $s_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $s_\mathcal{L} = s_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$, and that $s_{\qcs{L}\otimes \qcs{L'}} = s_{\qcs{L}} \, s_\qcs{L'}$
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.
We claim that $s_\qcs{L}$ is fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
\begin{align*}
S_G: \QCiso{G} &\to \Hh^0(\Weil{},\Hh^2(G(\bFq),\EEx))\\
\qcs{L} &\mapsto s_{\qcs{L}}.
\end{align*}

\begin{definition}\label{def:invisible}
Let $\qcs{L}$ be a quasicharacter sheaf on $G$. We say $\qcs{L}$ is \emph{visible} if $S_G(\qcs{L}) \in H^0(\Weil{},H^2(G(\bFq),\EEx))$ is trivial; otherwise, we say $\qcs{L}$ is \emph{invisible}.
\end{definition}

\subsection{Group cohomology}
%If $G$ is a connected and of finite type, then $\Hh^2(G(\bFq),\EEx)$ is trivial\todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible; by contrast, if $G$ is etale then $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$ may be non-trivial so $G$ may admit invisible quasicharacters sheaves. We will return to that case in Section~\ref{sec:etale}.
